### 2016-10-04 Paper_v7.tex
Ok

\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

\section*{Author Summary}
Tissue formation and remodeling requires a complex and dynamic balance of interactions between epithelial cells, which reside on the surface, and mesenchymal cells that reside in the tissue interior. During embryonic development, wound healing, and cancer, epithelial cells transform into a mesenchymal cell to form new types of tissues.
It is important to understand this process so that it can be controlled to generate beneficial effects and limit pathological differentiation.
Much research over the past 20 years has identified many different molecular species that are relevant, but these have mainly been studied one at a time.
In this study, we developed and implemented a novel computational strategy to interrogate all of the known players in this transformation process to identify which are the major bottlenecks.
We determined that NFATc1 and pSP1 are essential for promoting epithelial or mesenchymal differentiation, respectively.
We then predicted the existence of a partially transformed cell that exhibits both epithelial and mesenchymal characteristics.
We found this partial cell type develops a network of invasive but stunted vascular structures that may be a unique cell target for understanding cancer progression and angiogenesis.

@@

\section*{Introduction}
The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \citep{Thiery:2003il}. During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \citep{Stahl:2001pd}. EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \citep{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009). Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \citep{Zavadil:2005fu,Xu:2009oq}. Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding. Medici and coworkers recently identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \citep{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
Elucidating the master regulatory architecture controlling EMT therefore requires inclusion of these complex overlapping and non-binary behaviors.

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \citep{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \citep{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \citep{Vilar:2006gb}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \citep{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \citep{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
Recently, Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (JuPOETs) multiobjective optimization algorithm.
JuPOETs generated an ensemble of approximately 1400 models for analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.

@@

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} all of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test). The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT from a stable epithelial cell population required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving direct SNAIL/SLUG and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created the hypothetical YREG1 repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
In the absence of YREG1, the population of models failed to consistently to retain a stable epithelial state (Fig. \ref{fg:F4}D).
Conversely, YREG1 amplification revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fg:F4}C).
Elevated YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex, or TGF-$\beta$3 induced SMAD activation.
Taken together, low YREG1 expression was required for the maintenance of a stable epithelial phenotype that was simultaneously inducible across TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection, as seen in the training objectives.

@@

\section*{Discussion}
In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
The signaling architecture encoded in the model, which contained 74 molecular species interconnected by 169 interactions,
described the expression of 23 genes in response to growth factor stimulation.
This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the JuPOETs multiobjective optimization framework.
The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
The most novel hypothesis generated from the analysis was that cells could operate in a hybrid state defined by both epithelial and mesenchymal traits when stimulated simultaneously with TGF-$\beta$ and VEGF-A.
We tested this hypothesis in MCF10A and DLD1 cells stimulated with combinations of TGF-$\beta$ and VEGF-A.
As expected, in the presence of TGF-$\beta$ or VEGF-A alone, MCF10A and DLD1 cells were either mesenchymal or epithelial, respectively.
However, with both TGF-$\beta$ and VEGF-A, MCF10A and DLD1 cells exhibited a hybrid phenotype, having both epithelial and mesenchymal characteristics.
Furthermore, we found that functional traits such as tubulogenesis and ductal branching were different for cells in this hybrid phenotype.
Together, this study established a predictive model of EMT induction, determined that deterministic model ensembles could predict population heterogeneity, and proved the existence of a unique hybrid phenotype resulting from the simultaneous integration of extracellular growth factor signals.

>>>>

\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

\section*{Author Summary}
Tissue formation and remodeling requires a complex and dynamic balance of interactions between epithelial cells, which reside on the surface, and mesenchymal cells that reside in the tissue interior. During embryonic development, wound healing, and cancer, epithelial cells transform into a mesenchymal cell to form new types of tissues.
It is important to understand this process so that it can be controlled to generate beneficial effects and limit pathological differentiation.
Much research over the past 20 years has identified many different molecular species that are relevant, but these have mainly been studied one at a time.
In this study, we developed and implemented a novel computational strategy to interrogate the key players in this transformation process to identify which are the major bottlenecks.
We determined that NFATc1 and pSP1 are essential for promoting epithelial or mesenchymal differentiation, respectively.
We then predicted the existence of a partially transformed cell that exhibits both epithelial and mesenchymal characteristics.
We found this partial cell type develops a network of invasive but stunted vascular structures that may be a unique cell target for understanding cancer progression and angiogenesis.

@@

\section*{Introduction}
The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \citep{Thiery:2003il}. During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \citep{Stahl:2001pd}. EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \citep{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009). Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \citep{Zavadil:2005fu,Xu:2009oq}. Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding. Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \citep{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
Elucidating the master regulatory architecture controlling EMT therefore requires inclusion of these complex overlapping and non-binary behaviors.

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \citep{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \citep{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \citep{Vilar:2006gb}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \citep{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \citep{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.

In this study, we developed a family of mathematical models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We integrated a simple rule-based description of activity and gene expression regulation with traditional ordinary differential equation (ODE) modeling to describe an EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information, it also allowed to build a predictive yet compact model.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (JuPOETs) multiobjective optimization algorithm.
JuPOETs generated an ensemble of approximately 1400 models for analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated as ordinary differential equations (ODEs) augmented with rule-based descriptions of activity and gene expression regulation.
ODEs are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
Additionally, we used data generated in this study (Fig. \ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.

@@

The predictive power of the ensemble was tested using cross validation and by comparing simulations with data not used for model training.
In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} all of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test). The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT from a stable epithelial cell population required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving direct SNAIL/SLUG and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created the hypothetical YREG1 repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
In the absence of YREG1, the population of models failed to consistently to retain a stable epithelial state (Fig. \ref{fg:F4}D).
Conversely, YREG1 amplification revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fg:F4}C).
Elevated YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex, or TGF-$\beta$3 induced SMAD activation.
Taken together, low YREG1 expression was required for the maintenance of a stable epithelial phenotype that was simultaneously inducible across TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection.

%as seen in the training objectives.

@@

\section*{Discussion}
In this study, we developed a family of mathematical models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
The model, which contained 74 molecular species interconnected by 169 interactions,
described the expression of 23 genes in response to growth factor stimulation.
We estimated an ensemble of likely model parameters using the JuPOETs multiobjective optimization framework.
The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
The most novel hypothesis generated from the analysis was that cells could operate in a hybrid state defined by both epithelial and mesenchymal traits when stimulated simultaneously with TGF-$\beta$ and VEGF-A.
We tested this hypothesis in MCF10A and DLD1 cells stimulated with combinations of TGF-$\beta$ and VEGF-A.
As expected, in the presence of TGF-$\beta$ or VEGF-A alone, MCF10A and DLD1 cells were either mesenchymal or epithelial, respectively.
However, with both TGF-$\beta$ and VEGF-A, MCF10A and DLD1 cells exhibited a hybrid phenotype, having both epithelial and mesenchymal characteristics.
Furthermore, we found that functional traits such as tubulogenesis and ductal branching were different for cells in this hybrid phenotype.
Together, this study established a predictive model of EMT induction, determined that deterministic model ensembles could predict population heterogeneity, and proved the existence of a unique hybrid phenotype resulting from the simultaneous integration of extracellular growth factor signals.
